---
title: "Peak Bloom Prediction Demo"
author: "Nitul Singha"
date: "28/02/2023"
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, 
                      message = FALSE,
                      fig.align = 'center',
                      out.width = '80%')
```

## Instructions

In this analysis we demonstrate a very simple way of predicting the peak bloom date in the coming decade for all four locations required by the competition.
The models used here are very simple and are using only the historic data for these four locations, but no other information or covariates.
At the end of this document ((#appendix-rnoaa)[Appendix A]), we also demonstrate a simple way to get historic temperature data for the four locations via the `rnoaa` package.

For this demo analysis we are using methods from the _tidyverse_ of R packages.
They can be installed via

```{r, eval=FALSE}
#install.packages('tidyverse')
#install.packages('lubridate')

```

and then loaded via

```{r}
library(tidyverse)
library(lubridate)
```


## Loading the data
Variables

'location': location
'lat':Latitude
'long': Longitude
'alt':Altitude
'year': year of blossom
'bloom_date': date of blossom
 "month": month of blossom
"tempmax" : Avg. maximum temperature from december to february   
"tempmin" : Avg. minimum temperature from december to february   
"temp"    : Avg. temperature from december to february    
"dew"     : Avg. dew from december to february   
"humidity": Avg. humidity from december to february     
"precip"  : Avg. percipitation from december to february     
"snow"    : Avg. snow from december to february     
"windspeed": Avg. windspeed from december to february    
"moonphase" :  Avg. moonphase from december to february    
"solarradiation": Avg. solarradiation from december to february
"daytime": Avg. daytime (in sec) from december to february



```{r}

djf<-read.csv("data/final/Fwash_DJF.csv")%>% 
  bind_rows(read.csv("data/final/Fkyoto_DJF.csv")) %>% 
  bind_rows(read.csv("data/final/Fliestal_DJF.csv"))%>% 
  bind_rows(read.csv("data/final/FVancover_DJF.csv"))%>%
  mutate(month = month(parse_date_time(x = paste(year,bloom_doy), orders = "yj")))
         

```


## Predicting the peak bloom

A very simple method to predict peak bloom date in the future is to fit a least-squares line through the observed dates and extrapolate the regression function.
We only use data from 1881 to 2022.


data for dec,jan,Feb

```{r}

djf<-read.csv("data/final/Fwash_DJF.csv")%>% 
  bind_rows(read.csv("data/final/Fkyoto_DJF.csv")) %>% 
  bind_rows(read.csv("data/final/Fliestal_DJF.csv"))%>% 
  bind_rows(read.csv("data/final/FVancover_DJF.csv"))%>%
  mutate(month = month(parse_date_time(x = paste(year,bloom_doy), orders = "yj")))
         

```

## Fitting our first model

```{r}
m5 <- lm((bloom_doy) ~year+month +temp+humidity+windspeed+precip+daytime, data = djf)
summary(m5)
```
## Check for outlier and remove them from the dataset.
```{r}
djf <- djf %>%
  mutate(outlier=abs(rstandard(m5))>2.5)

djf<- djf %>% 
  filter(outlier == FALSE)
```

## Fit another model
We transform 'temp' to exp(temp) to achive normality

```{r}
m7 <- lm((bloom_doy) ~year+as.factor(month)+exp(temp)+humidity+windspeed, data = djf)
summary(m7)
```
## Checking residual distribution.
```{r}
res <- resid(m7)
qqnorm(res)

qqline(res)
```

### We can say that the residuals are normally distributed. As tha points are aligned with the 45 degree line.

## loading future data set to get the prediction.

```{r}
pred<-read.csv("data/final/Fwash_DJF_p.csv")%>% 
  bind_rows(read.csv("data/final/Fkyoto_DJF_p.csv")) %>% 
  bind_rows(read.csv("data/final/Flist_DJF_p.csv"))%>% 
  bind_rows(read.csv("data/final/FVancover_DJF_p.csv"))


```


## Fitting a logistic regression to predict the "month" for future dataset.
```{r}
temp<-djf%>%
  filter(month!=5)

temp$month<-ifelse(temp$month==3,1,0)

```


```{r}

m9 <- glm(month ~location+year+temp+daytime+moonphase+dew+humidity, data = temp, 
          family = binomial,
          na.action = na.exclude)

summary(m9)
```

```{r}
temp <- temp %>%
  mutate(p_hat = predict(m9, type="response"))

temp <- temp %>%
  mutate(Predicted = ifelse(p_hat > 0.4, "3", "4"))

library(gmodels)
CrossTable(temp$month, temp$Predicted,
           prop.r = FALSE, prop.c = FALSE, prop.t = FALSE, prop.chisq = FALSE)
```


## Fitting the model in future dataset.
```{r}

pred <- pred %>%
  mutate(p_hat = predict.glm(m9,pred,type="response"))

pred <- pred %>%
  mutate(month = ifelse(p_hat > 0.4, 3, 4))

pred$bloom_doy<-predict.lm(m7,newdata = pred)
pred$bloom_doy<-round(pred$bloom_doy,0)
```
```{r}
pred<-pred%>%
  select(c("location","year","bloom_doy"))

write.csv(pred, "data/final/final.csv", row.names=FALSE)


```
